---
title: "HotSpot_NLCD_ChiSq"
author: "James Tsalah"
date: "2024-01-22"
output:
  pdf_document: default
  html_document: default
---

# Load Packages & Prepare Data
## Packages
```{r results='hide', message=FALSE, warning=FALSE}
  require(sf)
  require(tidyverse)
  require(terra)
```

## Prepare Data
### File Preparation Version
```{r}
  

  # Optional - Read in Clipped NLCD
  hs_nlcd = rast(here::here('data', 'nlcd_2021', 
                            'nlcd_wgs84.tif'))
```
### Folder Preparation Version
```{r}
  # shapefile_folder <- here::here('ss', 'mc_shp')
  # shapefile_folder <- here::here('ss', 'shp')

  read_and_prepare_data <- function(shapefile_folder) {
    shapefiles <- list.files(shapefile_folder, pattern = "\\.shp$", full.names = TRUE)
    sf_list <- shapefiles %>% map(~st_read(.x, quiet = TRUE))
    hs <- bind_rows(sf_list)
    return(hs)
  }
  
  spring_folder <- here::here('SS_Runs', 'ss_1234rm_seasonal23_r125m','shp', 'Spring')
  summer_folder <- here::here('SS_Runs', 'ss_1234rm_seasonal23_r125m', 'shp', 'Summer')
  fall_folder <- here::here('SS_Runs', 'ss_1234rm_seasonal23_r125m','shp', 'Fall')
  
  hs_spring = read_and_prepare_data(spring_folder)
  hs_summer = read_and_prepare_data(summer_folder)
  hs_fall = read_and_prepare_data(fall_folder)
  
  # Optional - Read in Clipped NLCD
  hs_nlcd = rast(here::here('data', 'nlcd_2021', 
                            'nlcd_wgs84.tif'))
```

# Run HotSpot NLCD Analysis
## Setup HotSpot Analysis Dependencies
```{r}
  # Initialize an empty dataframe to store results
  local_proportions <- data.frame(PolygonID = integer(),
                                  Class = character(),
                                  Count = integer(),
                                  Proportion = numeric())
  
  # NLCD Legend Mapping
  nlcd_legend <- data.frame(
    Red = c(11, 12, 21, 22, 23, 24, 31, 41, 42, 43, 51, 52, 71, 72, 73, 74, 81, 82, 90, 95),
    Class = c("Open Water", "Perennial Ice/Snow", 
              "Developed, Open Space", "Developed, Low Intensity", "Developed, Medium Intensity", "Developed, High Intensity", 
              "Barren Land",
              "Deciduous Forest", "Evergreen Forest", "Mixed Forest", 
              "Dwarf Scrub", "Shrub/Scrub",
              "Herbaceous", "Sedge/Herbaceous", "Lichens", "Moss", 
              "Hay/Pasture", "Cultivated Crops",
              "Woody Wetlands", "Emergent Herbaceous Wetlands")
  )
```

## Calculate HotSpot NLCD Proportions
```{r}
  # Calculate Proportions Custom Function
  calculate_local_proportions <- function(hs, hs_nlcd, nlcd_legend) {
    local_proportions <- data.frame(PolygonID = integer(), 
                                    Class = character(), 
                                    Count = integer(), 
                                    Proportion = numeric())
    for (i in 1:nrow(hs)) {
      single_poly <- hs[i, ]
      clipped_raster <- crop(hs_nlcd, single_poly) %>% mask(single_poly)
      nlcd_val_single <- values(clipped_raster)
      prop_single <- na.omit(data.frame(Red = nlcd_val_single)) %>%
        group_by(Red) %>%
        summarise(Count = n()) %>%
        mutate(Proportion = (Count / sum(Count))) %>%
        left_join(nlcd_legend, by = "Red") %>%
        select(Class, Count, Proportion)
      prop_single$PolygonID <- i
      local_proportions <- rbind(local_proportions, prop_single)
    }
    return(local_proportions)
  }
  
  # Run Proportion Calculation per Season
  spring_local_proportions  = calculate_local_proportions(hs_spring, hs_nlcd, nlcd_legend)
  summer_local_proportions = calculate_local_proportions(hs_summer, hs_nlcd, nlcd_legend)
  fall_local_proportions    = calculate_local_proportions(hs_fall, hs_nlcd, nlcd_legend)
```

# Calculate Summary Statistics for HotSpot NLCD Analysis
## Prepare Study Proportions
```{r}
  nlcd_study_val = read.csv(here::here('data', 'nlcd_2021','NLCD_Study_Values.csv')) 
  nlcd_study_prop = data.frame(Class = nlcd_study_val$NLCD_Land,
                               Count = nlcd_study_val$Count)
  nlcd_study_prop <- nlcd_study_prop %>%
    mutate(StudyAreaProportion = (Count / sum(Count)))
```

# Calculate Summary Statistics Function
```{r}
  # Summary Statistics Custom Function
  summary_statistics <- function(local_proportions, nlcd_study_prop) {
    # First, compute summary statistics for each Class in local_proportions
    class_summary <- local_proportions %>%
      group_by(Class) %>%
      summarise(
        MeanProportion = mean(Proportion, na.rm = TRUE),
        MedianProportion = median(Proportion, na.rm = TRUE),
        MinProportion = min(Proportion, na.rm = TRUE),
        MaxProportion = max(Proportion, na.rm = TRUE),
        SDProportion = sd(Proportion, na.rm = TRUE)
      )
    # Join with class_summary
    class_summary <- left_join(class_summary, nlcd_study_prop, by = "Class")
    # Rearrange columns to make StudyAreaProportion the second column
    class_summary <- class_summary %>%
      select(Class, StudyAreaProportion, everything())
    return(class_summary)
  }
  
  # Run Summary Statistics
  spring_class_summary = summary_statistics(spring_local_proportions, nlcd_study_prop)
  summer_class_summary = summary_statistics(summer_local_proportions, nlcd_study_prop)
  fall_class_summary   = summary_statistics(fall_local_proportions, nlcd_study_prop)
```

# Calculate ChiSq for each Land Cover Type, per Season
## Run ChiSq Analysis per Season
```{r}
  # Calculate Class ChiSq Custom Function
  hs_nlcd_chisq <- function(mean_proportions, nlcd_study_prop) {
    # Function to standardize proportions
    standardize_proportion <- function(proportion, base) {
      count <- proportion * base
      if (count - floor(count) > 0.5) {
        return(ceiling(count))
      } else {
        return(floor(count))
      }
    }
  
    # Calculate base number
    base_number <- sum(nlcd_study_prop$Count)
  
    # Apply standardization and check the output
    std_counts_local <- mapply(standardize_proportion, 
                               mean_proportions$MeanProportion, 
                               MoreArgs = list(base = base_number))
    std_counts_nlcd <- mapply(standardize_proportion, 
                              nlcd_study_prop$StudyAreaProportion, 
                              MoreArgs = list(base = base_number))
  
    mean_proportions$STDCount <- std_counts_local
    nlcd_study_prop$STDCount <- std_counts_nlcd
  
    # Initialize results data frame
    chisqr_results <- data.frame(Class = character(), 
                                 ChiSquared = numeric(), 
                                 pvalue = numeric(),
                                 stringsAsFactors = FALSE)
  
    # Perform chi-square tests
    for (class in unique(mean_proportions$Class)) {
      hotspot_count <- mean(mean_proportions$STDCount[mean_proportions$Class == class])
      study_count <- nlcd_study_prop$STDCount[nlcd_study_prop$Class == class]
      
      chisq_test <- chisq.test(c(hotspot_count, study_count))
      
      chisqr_results <- rbind(chisqr_results, 
                              data.frame(Class = class, 
                                         ChiSquared = chisq_test$statistic,
                                         pvalue = chisq_test$p.value))
    }
  
    return(chisqr_results)
  }
  
  # Standardize hotspots to study area proportions
  base_number <- sum(nlcd_study_prop$Count) 
  
  # Run ChiSq Analysis
  spring_chisq = hs_nlcd_chisq(spring_class_summary, nlcd_study_prop)
  summer_chisq = hs_nlcd_chisq(summer_class_summary, nlcd_study_prop)
  fall_chisq   = hs_nlcd_chisq(fall_class_summary, nlcd_study_prop)
```

## View Seasonal ChiSq Results
Note: Mean Proportion is referring to the hotspots!
```{r}
print(spring_class_summary[,1:3], row.names = FALSE)
print(spring_class_summary[,c(1, 4:7)], row.names = FALSE)
print(as_tibble(spring_chisq), row.names = FALSE)
```

```{r}
print(summer_class_summary[,1:3], row.names = FALSE)
print(summer_class_summary[,c(1, 4:7)], row.names = FALSE)
print(as_tibble(summer_chisq), row.names = FALSE)
```

```{r}
print(fall_class_summary[,1:3], row.names = FALSE)
print(fall_class_summary[,c(1, 4:7)], row.names = FALSE)
print(as_tibble(fall_chisq), row.names = FALSE)
```